Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} Model System

### 2014-05-28 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex

\documentclass[a4paper,11pt]{article}

\title{Authors' Reply to Reviewers' Comments on\\ \itshape{``Evaluating the Performance and Energy Efficiency of the COSMO-ART Model System''}}

\author{Joseph~Charles, William~Sawyer, Manuel~F.~Dolz, Sandra~Catal\'an}

\usepackage[left=2.5cm,right=2.2cm,top=3cm,bottom=3cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage{color}
\usepackage{graphicx}
\usepackage{url}
\usepackage[urlcolor=blue,colorlinks=true]{hyperref}
\usepackage{ifthen}

%\newcommand{\xavi}[1]{\textcolor{blue}{#1}}
%\newcommand{\xavi}[1]{{#1}}
%\newcommand{\xaviNote}[1]{\footnote{\textcolor{blue}{#1}}}

\newcounter{question}
\setcounter{question}{1}

%\newcommand{\reviewerSay}{\vspace{0.2cm} {\bf \noindent \arabic{question}} \fbox{{\sc \textcolor{red}{Reviewer:}}} \medskip \rm}
\newcommand{\reviewerSay}{\vspace{0.2cm} {\bf \noindent } {\bf Reviewer:} \medskip \par}
\newcommand{\editorialSay}{\vspace{0.2cm} {\bf \noindent \arabic{question}.} {\bf Editorial:} \medskip \par}
%\newcommand{\weReply}{\vspace{0.25cm} {\bf \noindent \arabic{question}} \addtocounter{question}{1} \fbox{{\sc \textcolor{blue}{Authors:}}} \medskip \em\par}
\newcommand{\weReply}{\vspace{0.25cm} {\bf \noindent } \addtocounter{question}{1} \textbf{\emph{Authors:}} \medskip \em\par}
\newenvironment{weSay}{\bigskip \weReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}
\newenvironment{weSayToEditorial}{\bigskip \weReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}

%\newcommand{\reviewerSay}{\paragraph*{(\arabic{question}) Reviewer: }}
%\newcommand{\authorsReply}{\paragraph*{\emph{(\arabic{question}) \addtocounter{question}{1} Authors: }}}
%\newenvironment{weSay}{\vspace{-3ex} \authorsReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}

\begin{document}

\maketitle

\noindent 
We would like to thank the Anonymous Reviewers
for providing us with a prompt and insightful review. 
We take this as an indication that there was genuine interest in the paper,
which is always a compliment regardless of the outcome of the review.

We addressed all the suggestions and recommendations from the reviewers to improve
the quality and flow of the paper. Below, we discuss the comments and suggestions
from the reviewers in detail.

\newpage
% ------------------------------------------------------------------------------------------------------------------------------------------------
\section*{Reviewer \#1}

\reviewerSay
The authors perform an energy performance and efficiency analysis of the COSMO-ART model system,
a complex weather prediction and climate model, focused on the study of aerosol concentrations in the
atmosphere. They present the model, the software and the hardware setup, including two metering
frameworks and perform out-of-band (i.e. measurement from external tools) power draw measurement
for three different setups. They use the experiments to perform two different analyses: energy 
efficiency for Sandy Bridge versus Ivy Bridge, and a power profile to compare polling and blocking 
MPI policies that also show the heavy energy consuming phases of the COSMO-ART model system.

The paper is well written and is easy to follow, and, in spite of the idea not being completely new, 
it offers a good insight about where are the most energy demanding phases and how the different 
policies on MPI affect them for the second experimentation part. The first experimentation part, 
besides being a common study, gives an interesting point of view about how energy efficient does 
not necessarily means time efficient.

%%% *** Manuel/Sandra ***
\begin{weSay}
The authors thank the reviewer for his/her general view of the paper.
\end{weSay}

Some suggestions regarding the paper:

- MONCH and PILATUS are presented as Ivy Bridge and Sandy Bridge clusters respectively in Section 4.1, 
and afterwards in section 5, they are referred as Ivy Bridge and Xeon E5 respectively. Apart from not 
being consistent, both processor families are Xeon E5 thus confusing a bit on what cluster is 
being referred. This also affects Table 1.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- It would be easier to compare values in Figure 4 if both energy ETSs and TTSs where together. 
Right now you show ETS TTS ETS TTS, and ETS ETS TTS TTS would be easier to understand.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- In multiple places: "power dissipation", "power draw". I assumed it this way, if it is not 
I suggest to explain it better.

%%% *** Manuel ***
\begin{weSay}
blabla
\end{weSay}

\newpage
\section*{Reviewer \#2}

\reviewerSay
The paper investigates the energy footprint of a weather forecasting model on various HPC platforms. 
As such the paper provides the overall energy footprint covering the overall run. In other words the 
granularity is very coarse and is at the application level. Although this will be interesting to the 
COSMO-ART community, the results include some good details on traces  on the
evaluation of polling vs blocking MPI strategies.

It could have been very beneficial to describe the nature of the application in detail instead of 
citing them. I assume that authors have resorted to that technique for the reasons of brevity  - yet 
I beleive that such a detailed description is essential to understand the behaviour of the application.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

With missing crucial details of the application and in the abseence of any results towards various 
problem sizes or domains of cases, it is very difficult to do any post-analysis on the overall 
behaviour of the application - for example its runtime scalability (both weak and strong) vs energy 
scalability etc.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In summary, although the paper is very interesting, the lack or depth of information in there severely 
limits the re-usable knowledge contributed by the paper.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

\newpage
\section*{Reviewer \#2}

The authors describe the extended atmospheric model COSMO-ART. They present
performance and energy measurements on a selection of HPC systems. There
is no new methodology presented in this paper.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

Related work is weak. I don't see how [20] and [11] are related to the content
of this paper (other algorithms, other architectures, no comparison of the
energy measurement infrastructure). I suggest to focus more on infrastructure
and tool support for system level power measurement in HPC environments in
the related work section. This would give you the opportunity to argue that
the 1 Hz resolution in one of your test systems is actually state-of-the-art.

%%% *** Will/Joseph ***
\begin{weSay}
blabla
\end{weSay}

The hardware description is a bit confusing.
- why does the NEC system have 10 racks for 312 nodes if one rack houses 52
     nodes?

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- the "v2" is missing from the Ivy Bridge processor's name string

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- sentence structure: "Each node is comprises two..."

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- what do you mean by "second-generation processor" regarding Sandy Bridge?
     SB is sometimes referred to as 2nd gen Core-i* processor but this does not
     apply to the Xeon line

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- the claim that SB is known to consume more power is wrong. The TDP classes
     are the same for IB and SB (e.g. E5-2670 and E5-2670v2 are both 115W parts)
     It presumably is less efficient, but even such a claim would need a reference

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- It should be added if the Turbo Boost feature is enabled or disabled, this
     seems to be important for the discussion of the TTS results.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

- It should be added which power saving mechanisms are used (cpufreq
     governor, available c-states)

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

- the second system is described as 16 cores x 42 nodes = 672 cores total. In
     4.3 it is stated that 1344 cores were used (Could be HyperThreading, however
     it is also stated that HT was not considered. Also, 16 MPI ranks per node
     indicate 84 nodes)

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

In 4.2.1
A 5 minute sampling interval seems useless to me. It would be nice to know if
this is actually an instantaneous value every 5 minutes (as I read it) or the
correct average of the past 5 minutes. Figure 5 suggests variation of the power
consumption in the 10\% range. Thus instantaneous values would have a high margin
for error and the 4,6, or 9 samples you use to calculate the average would be
too few to estimate a reliable average.

%%% *** Will/Joseph ***
\begin{weSay}
blabla
\end{weSay}

In 4.2.2
There is no information about the accuracy

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

In 4.3
Only -O3 is not appropriate for the Sandy bridge and Ivy Bridge system. At
least -mavx should be added. It looks like some low hanging fruit is reserved
to reach the targeted 5x improvement

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In 5.1
The blocking mode does not seem to be blocking. As there are no other
processes/threads available (as many MPI ranks as cores per node) the waiting
ranks will repeatedly call sched\_yield() only to be picked again by the OS
scheduler. This also explains the small benefit as no idle states are used.
You should check the CPU utilization. I guess it is 100\% on all cores in both
modes. If it is not, you should check with powertop which idle states are used.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

In 5.2
What do you mean by "it aims at computing speed regardless to energy consumption"?
Does it mean it uses Turbo? Please clarify what exactly is different than on the
IB system.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

It is confusing to compare "Ivy Bridge EP" against "Xeon E5" in the result
discussion, as the Ivy Bridge processor is also called Xeon E5 (v2) as well.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

Since the SB and IB measurements use different amounts of cores it is essential to
know if the application scales linearly with the core count. If the application
is scalable 1020 cores at 2.2 GHz should be faster than 672 cores at 2.6 GHz.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In 5.3
Only 5\% energy reduction when blocked shows that the processors are not idle.
As mentioned earlier, you should add information if power management is
actually used while the processors are waiting.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

\end{document}

### 2014-05-28 trunk/paper_enahpc14/Introduction.tex
updates a few things according to the reviewers' suggestions

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid-20$^{th}$  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  determining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  heat-trapping greenhouse  gases and
have  shorter  atmospheric lifetimes,  aerosols  significantly modify  the
global   radiation   budget   \cite{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
modifying cloud  properties,  cloud  fraction and  surface
albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
considerable  progress in  global aerosol  modelling \cite{Mann-2013}
and  measurement-based  assessments,  large  uncertainties  remain  in
current  estimates  of  aerosol radiative  forcing  \cite{Myhre-2013,
IPCC-2013, Lee-2013,  Sherwood-2013,  Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale modelling system is coupled  with the operational  weather forecast
model  \textsc{Cosmo}  \cite{Baldauf-2011},  jointly developed and used by  a
consortium of European weather centers, as well as utilized in a climate version
by the wider research community.

The extended  atmospheric model \cosmoart  is com\-put\-ationally
much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
additional  tracers and  processes have  to be  considered.   Thus the
model  is currently  severely limited  in terms  of applicability  and
expensive in terms of energy consumption.  Although \textsc{Cosmo} has
recently been ported to GPUs \cite{Gysi-2014, Lapillonne-2014} 
within
the framework of the  High Performance and High Productivity Computing
(HP2C) Initiative  \cite{HP2C} to optimize it  for computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The  efficiency of \cosmoart
is being  addressed in the  EU Exa2Green project  \cite{EXA2GREEN} to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This  work is  organized as  follows: in  Section~\ref{sec:1} we  give an
overview  of  related  work,   then  in  Section~\ref{sec:2}  we  briefly
introduce \cosmoart  and specify its  technical setup related
to  the investigated  performance and  energy evaluation  methods.  In
Section~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking study.  Section~\ref{sec:4}  presents performance and energy
requirements  of the  baseline on  these architectures  and highlights
areas where improvements will be necessary for the subsequent baseline
refactoring.   Finally,  we  conclude in Section~\ref{concl} with some 
implications for the Exa2Green project and give an outlook for future
research.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate changes  since  the mid-20$^{th}$  Century, one  factor
overwhelmingly  affects the  uncertainty in  determining human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not  considered heat-trapping greenhouse  gases and
have shorter atmospheric  lifetimes, aerosols significantly modify the
global   radiation   budget   \cite{IPCC-2013}.    Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
modifying cloud properties, cloud fraction and surface albedo, causing
a negative indirect  radiative forcing.  Despite considerable progress
in  global aerosol  modelling  \cite{Mann-2013} and  measurement-based
assessments,  large  uncertainties  remain  in  current  estimates  of
aerosol  radiative   forcing  \cite{IPCC-2013,  Lee-2013,  Myhre-2013,
Randles-2013, Rosenfeld-2013, Sherwood-2013, Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
scale  modelling  system  is  coupled  with  the  operational  weather
forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
and  used by  a consortium  of European  weather centers,  as  well as
utilised in a climate version by the wider research community.

The extended  atmospheric model \cosmoart  is com\-put\-ationally much
more demanding than \textsc{Cosmo}  since a large number of additional
tracers  and processes  have  to  be considered.   Thus  the model  is
currently severely limited in  terms of applicability and expensive in
terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
framework  of the  High  Performance and  High Productivity  Computing
(HP2C)  Initiative \cite{HP2C}  to optimise  it for  computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The efficiency  of \cosmoart is being
addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
prototype code, which  provides an energy efficiency of  at least five
times  the baseline  value.  Such  an implementation  would  allow the
community to  investigate critical questions at  higher resolution and
over longer periods, at reduced cost to the environment.

This work is  organised as follows: in Section~\ref{sec:1}  we give an
overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
introduce  \cosmoart and specify  its technical  setup related  to the
investigated   performance   and   energy  evaluation   methods.    In
Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
energy  requirements  of  the  baseline  on  these  architectures  and
highlights  areas  where  improvements   will  be  necessary  for  the
subsequent   baseline   refactoring.     Finally,   we   conclude   in
Section~\ref{concl} with  some implications for  the Exa2Green project
and give an outlook for future research.

### 2014-05-28 trunk/paper_enahpc14/Sec3:Environment-setup.tex
updates a few things according to the reviewers' suggestions

\begin{itemize}
\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed of
  312 compute nodes. 
  % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
  % Scientific Computing (PASC) project~\cite{PASC}.
  Each node is  comprised of two Intel Ivy  Bridge EP E5-2660 ten-core
  processors  operating  at 2.2\,GHz,  equipped  with  32\,GB of  DDR3
  1600\,MHz RAM  and connected via InfiniBand Mellanox  SX6036 and FDR
  switches. For our  experiments, a full rack constituted  of 52 nodes
  was used. 
  % \monch is  slated to stay  in service without hardware  upgrade for
  % the  duration  of  the  Exa2Green project,  allowing  an  identical
  % configuration  to  be  recreated  for  future  assessments  of  the
  % baseline.

\item \pilat is  a dual-socket eight-core Intel Sandy  Bridge EP based
  cluster at CSCS composed of 42 compute nodes.  
  % and used as Piz Daint  pre-post processing cluster The 2 login and
  % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
  % E5-Series  DALCO  r2264i4t   2U  scalable  compute  modules,  each
  % containing 4 compute nodes based on two ...
  Each node  is comprises two Intel Xeon  E5-2670 processors operating
  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected
  via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
  experiments, a full  rack constituted of 42 compute  nodes was used.
  \pilat   is   based   on   Intel's   second-generation   processors,
  conventional in  HPC systems  and known to  be more  power consuming
  than its Ivy Bridge successor.

>>>>

\begin{itemize}
\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed by
  376  compute  nodes  grouped  in  312  standard  compute  nodes,  24
  large-memory compute nodes,  24 huge-memory nodes, 16 administration
  nodes and 8 servers to provide I/O to Lustre.
  % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
  % Scientific  Computing (PASC)  project~\cite{PASC}.   Each standard
  compute  node is  comprised of  two  Intel Ivy  Bridge EP  E5-2660v2
  ten-core processors  operating at 2.2\,GHz, equipped  with 32\,GB of
  DDR3 1600\,MHz RAM and  connected via InfiniBand Mellanox SX6036 and
  FDR  switches.   For  our   experiments,  a  full  rack  constituted
  exclusively of 52 nodes was used.
  % \monch is  slated to stay  in service without hardware  upgrade for
  % the  duration  of  the  Exa2Green project,  allowing  an  identical
  % configuration  to  be  recreated  for  future  assessments  of  the
  % baseline.

\item \pilat is  a dual-socket eight-core Intel Sandy  Bridge EP based
  cluster at CSCS composed of 42 compute nodes.  
  % and used as Piz Daint pre-post processing cluster. The 2 login and
  % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
  % E5-Series  DALCO  r2264i4t   2U  scalable  compute  modules,  each
  % containing 4 compute nodes based on two ...
  Each node  is comprised two Intel Xeon  E5-2670 processors operating
  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected
  via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
  experiments, a full  rack constituted of 42 compute  nodes was used.
  \pilat   is   based   on   Intel's   second-generation   processors,
  conventional in  HPC systems  and known to  be more  power consuming
  than its Ivy Bridge successor.
